Fast de novo transposable element annotation

### 2015-09-14 manuscript/JCB/bmc_article.tex
Manuscript update

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
\section*{Background}
Transposable elements, or TEs, are genomic sequences that have at some
point had the capacity to insert copies of themselves into other
genomic locations, resulting in homologous families of sequences
spread across the genome.  TEs are present in almost every higher
order genome, covering as much as 45\% of the human genome and 90\% of
the maize genome \cite{Venter:2001p92,SanMiguel:1996wa}.  TEs have
proved an important source of data in a number of studies
\cite{Arndt:2005p279,Karro:2008p123,Mugal:2009p581,Hardison:2003p97},
but given their prevalence, it is just important for those studying
the genome to have TEs masked out -- their bases replaced by N to
allow for easy identification and filtering.  Failure to filter can
reek havoc with genomic analysis tools.  For example. if not filtered,
TEs can trigger huge numbers of false positives automated gene finding
tool \cite{Jiang:2013jt}, as well as inflate tool runtime.

The most well known tools for repeat identification is RepeatMasker,
with the new nHMMER showing promise as well)
\cite{RepeatMaskerOpen:XkNxXSd7,Wheeler:2013gj}.  Both work on the
principle of a library-based search.  That is, to identify
transposable elements to a known family, the tools require some
description of sequences in the family (e.g. a ancestral sequence for
BLASTing, or a profile HMM description) that is used as the basis for
identification.  But, much like we ask how the snow plow driver gets
to work, or how the author of a dictionary authors look up the spelling of a word
\cite{Pratchett:uw}, we must ask how the libraries on which the tools
are based are initially compiled.  Library-based tools are useless for the
discovery of new families, and cannot be applied to a newly sequenced
genome for which such a library has not been compiled.

Within mammalian species we can largely rely on homology
relationships to port libraries across species.  A library description
of a human ALU sequence can likely be used for the initial annotation
of a newly sequenced primate genome, and many human LINEs will extend to
rodent, or even more distant, mammalian species.  This does not hold
so well in plants: in many cases TE composition of a given plant organism
is species-specific.  For example, a rice-based TE library will only identify 25\%
of the TEs in the maize genome \cite{Jiang:2013jt}.

To solve this problem we turn to {\it de novo} TE identification
tools: tools identifying TEs though a genome using only the genome
sequence information.  A number of such tools are discussed in the
literature, operating on a number of different principles.  Tools such
as RECON and PILER are based on
self-alignment, using WU-BLAST and LASTZ for the alignment
\cite{Bao:2002,Edgar:2005p2365,Lopez:2003td,Harris:2007uf}.  RECON show good sensitivity but is
computationally intensive and infeasible for use on whole genomes
(requiring 60 hours for 18Mb rice genome int 2013), while
PILER achieves a good runtime with very low sensitivity
\cite{Jiang:2013jt}.  ReAS \cite{Li:2005he} and RepeatScout
\cite{Price:2005p1247} are based on $k$-mer searches, with the
earlier showing less sensitivity than RECON \cite{Jiang:2013jt}.
RepeatGluer \cite{Pevzner:2004p3157,Zhi:2006p3199} is
based on a variation of DeBrujin graphs, which allows for a
decomposition of TE families into domains, but is very, very
computationally expensive.  In Saha {\it et al.} the authors perform
an extensive comparison of the tools, and conclude that RepeatScout is
the best tool overall for assembled genomes, while ReAS the best when
dealing with unassembled sequence fragments \cite{Saha:2008dm}.

@@

While it is notoriously difficult to mathematically model transposable
elements \cite{Bao:2002}, elementary repeats are more conducive to a
formal description.  For a given genome, a nucleotide sequence $r$ is
an elementary repeat if:
\begin{enumerate}
\item It is of at least length $l$.  (The length requirement.)
\item At least $f$ exact copies of $r$ appear in the genome.  (The
  frequency requirement.)
\item There is no proper substring of $r$ that is of at least length
  $l$ and appears in the genome independently of $r$. (The minimality
  requirement.)
\item There is no string that contains $r$, appears in the genome
  the same number of times as $r$, and satisfies conditions 1-3.  (The
  maximality requirement.)
\end{enumerate}
The first two criteria are straight-forward, allowing us to designate
length and frequency thresholds for our model.  The minimality
requirement ensures that an elementary repeat does not contain
anything that could be viewed as an independent elementary repeat --
that it is the ``smallest building block'' that we can find.  While
the maximality requirement ensures that it is not tied to some other, larger,
elementary repeat.

Having proposed this definition, Zheng and Lonardi developed an
identification algorithm that had a runtime quadratic in the genome
size -- so not of practical use on whole-genome analysis.  This was
refined to linear time by He and also by Huo {\it et al.}
\cite{He:2006gpa,Huo:2009hoa} based on variations of suffix tree
approaches, but are limited
in their ability to handle sequence variation.  As transposable
elements naturally suffer from copy mistakes and accumulate
instance-specific base substitutions over time, this is a significant
limitation.

\subsection*{RAIDER}
\label{RAIDERSec}
It was with the objective of creating a linear time identification
algorithm that could handle variation through use of
PatternHunter-like spaced seeds that we developed the prototype RIADER
\cite{Li:2004wl}.  A rough implementation was first presented in
Figueroa {\it et al.}, with more details in the Figueroa masters
thesis \cite{Figueroa:2014uk,Figueroa:2013cz}.  Not itself fully able
to handle the spaced seed component, RAIDER was built along an
alternate, but equivalent definition of elementary repeats based on
$l$-mers (sequences of length $l$).  Specifically, it was observed
that the minimality condition could be rewritten as: 
There is no $l$-mer contained with $r$ that appears in the genome
more times than $r$.  From there we make our four core observations that form
the basis for the RAIDER algorithm:
\begin{enumerate}
\item An $l$-mer cannot belong to two different elementary repeats.
\item Any $l$-mer in the genome that occurs $f$ or more times is
  either an elementary repeat or belongs to one.
\item Any two $l$-mers belonging to an elementary repeat must appear
  the same number of times in the genome. 
\item If two sequences in the genome that are {\it maximally
  identical} (that is, cannot be extended in either direction and still
  be identical), these sequences cannot belong to a larger
  elementary repeat.
\end{enumerate}
By ``belong'' we mean ``is a substring of''; the reason for the
terminology will be explained shortly.  For discussion and proof, see
the Figueroa Thesis \cite{Figueroa:2013cz}.

@@

Results of the preliminary implementation were promising.  Based on
the Saha {\it et al.} analysis putting RepeatScout as the top {\it de
  novo} search tool, comparisons were made against that tool.  RAIDER result
quality was close, but slightly behind -- with runtime
order of magnitudes better.  On human chromosome 22 we saw a
$12\times$ speedup from RepeatScout to RAIDER (2344 seconds to 192 seconds), while coverage
of the RepBase  \cite{Jurka:2005bl} ancestral sequence actually improved (77\% to 84\%), while on mouse chromosome 19 we saw 
the same speedup with a significant drop in coverage (53\% to 30\%).  On
the full human genome RAIDER ran in $6.3$ hours, while RepeatScout was
unable to complete it run.  For details, see Figueroa {\it et al.} \cite{Figueroa:2014uk}.

@@

PatternHunter, first proposed as a very successful augmentation to
BLAST \cite{Li:2004wl,Altschul:1997p843}, is based on the notion of {\it
  spaced seeds}.  In the context of BLAST, instead of looking for a
contiguous substring match to trigger the alignment extension,
PatternHunter uses a more refined match pattern.  For example, the
seed 1111110111111 would specify that we were looking for an exact
match of two length-six substrings separated by a single base (which
might or might not match).  In the toy seed $s=110101$, we would say the
strings $AACACA$ and $AAGACA$ match because all miss-matched positions correspond to a 
0 match.  The string $AACACA$ does not math $TACACA$ (with respect to
$s$), as the single miss-match does not correspond to the 1.  Nor does
$AACACA$ match any substring of $TTAACTCA$, though it does match a
substring of $T\mathbf{AAGATA}C$.

The incorporation of spaced seeds into BLAST lead to significant
improvements at virtually no cost in runtime \cite{Li:2004wl}.
In our context, we will use them to allow limited
miss-matches between instances of ``the same'' elementary repeat.  The
preliminary version of RAIDER from Figueroa {\it et al.} made some
attempt to incorporate this.  It was successful enough to provide a
proof-of-concept, but a significant amount of work was left to be
done to allow for true space seed usage.

In this paper we present phRAIDER (PatternHunter-based RAIDER), a fast
tool for the identification and making of transposible elements in
both assembled and unassembled genomes. Code  may be obtained [NEED WEB ADDRESS].  In our Methods section
we present the new model that allows us to extend RAIDER to correctly
use spaced seeds, and briefly outline the algorithm (with more details
provided in the supplementary materials), with a analysis of phRAIDER
performance in our Results.

\section*{Methods}

In order to allow the use of a spaced seed strategy, we will need to
define a new model of elementary repeats that will accommodate it.  In
the following we will first outline our theoretical framework
(described in detail in the Supplementary Materials), and then review
the phRAIDER algorithm.

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
\section*{Background}
Transposable Elements (TEs) are genomic sequences that have at some
point had the capacity to insert copies of themselves into other
genomic locations, resulting in homologous families of sequences
spread across the genome.  Present in almost every higher
order genome (covering as much as 45\% of the human genome and 90\% of
the maize genome \cite{Venter:2001p92,SanMiguel:1996wa}), TEs have
proved an important source of data in numerous studies of genomic structure (e.g.
\cite{Arndt:2005p279,Karro:2008p123,Mugal:2009p581,Hardison:2003p97}).
But given their prevalence, it is important for those studying other aspect
the genome to have TEs masked out -- their bases replaced by N to
allow for easy identification and filtering.  Failure to filter can
reek havoc with genomic analysis tools.  For example,
TEs can trigger huge numbers of false positives automated gene finding
tool \cite{Jiang:2013jt}, as well as inflate tool runtime.

The best tools for repeat identification are RepeatMasker and nHMMER
\cite{RepeatMaskerOpen:XkNxXSd7,Wheeler:2013gj}, but both employ a
library-based search strategy.  They require some pre-compiled
description of sequences in the family (e.g. a ancestral sequence for
BLASTing, or a profile HMM) that is used as the basis for
identification of new member of the family.  But, much like we ask how
the snow plow driver gets to work \cite{Pratchett:uw}, we must ask how
these libraries are based are initially compiled.  Library-based tools
are useless for the discovery of new families, and cannot be applied
to a newly sequenced genome for which such a library has not been
compiled.

Within mammalian species we can largely rely on homology relationships
to port libraries across species.  This does not hold
so well in plants: in many cases TE composition of a given plant
organism is species-specific.  For example, a rice-based TE library
will only identify 25\% of the TEs in the maize genome
\cite{Jiang:2013jt}.

To solve this problem we turn to {\it de novo} TE identification
tools, identifying TEs using only the genome
sequence information.  A number of such tools are discussed in the
literature.   RECON and PILER are based on
self-alignment, using WU-BLAST and LASTZ for the alignment
\cite{Bao:2002,Edgar:2005p2365,Lopez:2003td,Harris:2007uf}.  RECON show good sensitivity but is
computationally intensive and infeasible for use on whole genomes
(requiring 60 hours for 18Mb rice genome in a 2013 study), while
PILER achieves a good runtime with very low sensitivity
\cite{Jiang:2013jt}.  ReAS \cite{Li:2005he} and RepeatScout
\cite{Price:2005p1247} are based on $k$-mer searches, with the
earlier showing less sensitivity than RECON \cite{Jiang:2013jt}.
RepeatGluer \cite{Pevzner:2004p3157,Zhi:2006p3199} is
based on a variation of DeBrujin graphs, which allows for a
decomposition of TE families into domains, but is very, very
computationally expensive.  In Saha {\it et al.} the authors perform
an extensive comparison of the tools, and conclude that RepeatScout is
the best tool overall for assembled genomes, while ReAS the best when
dealing with unassembled sequence fragments \cite{Saha:2008dm}.

@@

While it is notoriously difficult to mathematically model transposable
elements \cite{Bao:2002}, elementary repeats are more conducive to a
formal description.  For a given genome, a nucleotide sequence $r$ is
an elementary repeat if: {\bf (1)} It is of at least length $l$ (the
length requirement); {\bf (2)} there are at least $f$ copies of $r$
appear (the frequency requirement) {\bf (3)} there is no proper
substring of $r$ of length $\leq l$ that appears in the genome
independently of $r$ (the minimality requirement); {\bf (4)} $r$ is a
maximal string w.r.t (1-3) (the maximality requirement).  Having
proposed this definition, Zheng and Lonardi developed an
identification algorithm that had a runtime quadratic in the genome
size. This was refined to linear time by He and also by Huo {\it et
  al.}  \cite{He:2006gpa,Huo:2009hoa} based on variations of suffix
tree approaches, but are limited in their ability to handle sequence
variation.  As we are looking a genome size inputs, and TEs
transposable elements naturally suffer from copy mistakes and
accumulate instance-specific base substitutions over time, this is a
significant limitation.

\subsection*{RAIDER}
\label{RAIDERSec}
It was with the objective of creating a linear time identification
algorithm that could handle variation through use of
PatternHunter-like spaced seeds that we developed the prototype RIADER
\cite{Li:2004wl}.  A rough implementation was first presented in
Figueroa {\it et al.}, with more details in the Figueroa masters
thesis \cite{Figueroa:2014uk,Figueroa:2013cz}.  RAIDER was built along an
alternate, but equivalent definition of elementary repeats based on
$l$-mers (sequences of length $l$).  Specifically, it was observed
that the minimality condition could be rewritten as: 
There is no $l$-mer contained within elementary repeat $r$ that appears in the genome
more times than $r$.  From there we make four core observations that form
the basis for the RAIDER algorithm:
\begin{enumerate}
\item An $l$-mer cannot belong to two different elementary repeats.
\item Any $l$-mer in the genome that occurs $f$ or more times is
  either an elementary repeat or belongs to one.
\item Any two $l$-mers belonging to an elementary repeat must appear
  the same number of times in the genome. 
\item If two sequences in the genome that are {\it maximally
  identical} (that is, cannot be extended in either direction and still
  be identical), these sequences cannot belong to a larger
  elementary repeat.
\end{enumerate}
By ``belong'' we mean ``is a substring of''; the reason for the
terminology will be explained shortly.  For discussion and proof, see
the Figueroa Thesis \cite{Figueroa:2013cz}.

@@

Results of the preliminary implementation were promising. On human
chromosome 22 we saw a $12\times$ speedup over RepeatScout to RAIDER
(2344 seconds to 192 seconds), while coverage of the RepBase
\cite{Jurka:2005bl} ancestral sequence improved (77\% to 84\%), while
on mouse chromosome 19 we saw the same speedup with a significant drop
in coverage (53\% to 30\%).  On the full human genome RAIDER ran in
$6.3$ hours, while RepeatScout was unable to complete it run.  For
details, see Figueroa {\it et al.} \cite{Figueroa:2014uk}.

@@

PatternHunter, a very successful augmentation to
BLAST \cite{Li:2004wl,Altschul:1997p843}, is based on the notion of {\it
  spaced seeds}: improving the sensitivity of string matching based
algorthms by allowing wild-cards in the match.  That is, instead of
requiring two strings matching in 12 consecutive characters, we might
instead require two six-character exact matches seperated by one base
(represented by the {\it seed pattern} 11111101111111), or perhaps
three consecutive four-character exact matches sperated by two bases
each (1111001111001111). It has been demonstrated that certain seed
patterns can idunce significant imporvements in BLAST sensitivity with
out time penalty, though what makes a good pattern is not well understood.

RAIDER was designed with the intent of employing the spaced seed
strategy, but this was only implemented heuristically for the Figuera
{\it et al.} paper \cite{Figueroa:2013cz} -- serving primarily as a
proof-of-concept.  Since its release we have develoed a formal model
of elementary repeats that incorporated spaced seeds, and from that
developed phRAIDER (PatternHunter-based RAIDER).  phRAIDER is a fast
tool for the identification and making of transposible elements in
both assembled and unassembled genomes aht outpreforms RptScout and
other establihsed tools. Code is free available under the Gnu GPL
lisence (v. 3) and  may be obtained [NEED WEB ADDRESS].  In our Methods section
we present the new model that allows us to extend RAIDER to correctly
use spaced seeds, and briefly outline the algorithm (with more details
provided in the supplementary materials), with a analysis of phRAIDER
performance in our Results.

\section*{Model}

Our goal is to redefine the concept of transposbile elements to
accomodate a spaced seed strategy.  In the following section we will describe our
identification algorithm, and then describe the success in using it to
mase transposable elements. But we will start here with a brief
outline of our theoretical model (with a more detailed description in
the appendix).

@@

\section*{Appendix}
In order to allow the use of a spaced seed strategy, we will need to
define a new model of elementary repeats that will accommodate it.  In
the following we will first outline our theoretical framework
(described in detail in the Supplementary Materials), and then review
the phRAIDER algorithm.

### 2015-09-14 manuscript/JCB/manuscript.tex
Manuscript update

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
\section*{Background}
Transposable Elements (TEs) are genomic sequences that have at some
point had the capacity to insert copies of themselves into other
genomic locations, resulting in homologous families of sequences
spread across the genome.  Present in almost every higher
order genome (covering as much as 45\% of the human genome and 90\% of
the maize genome \cite{Venter:2001p92,SanMiguel:1996wa}), TEs have
proved an important source of data in numerous studies of genomic structure (e.g.
\cite{Arndt:2005p279,Karro:2008p123,Mugal:2009p581,Hardison:2003p97}).
But given their prevalence, it is important for those studying other aspect
the genome to have TEs masked out -- their bases replaced by N to
allow for easy identification and filtering.  Failure to filter can
reek havoc with genomic analysis tools.  For example,
TEs can trigger huge numbers of false positives automated gene finding
tool \cite{Jiang:2013jt}, as well as inflate tool runtime.

The best tools for repeat identification are RepeatMasker and nHMMER
\cite{RepeatMaskerOpen:XkNxXSd7,Wheeler:2013gj}, but both employ a
library-based search strategy.  They require some pre-compiled
description of sequences in the family (e.g. a ancestral sequence for
BLASTing, or a profile HMM) that is used as the basis for
identification of new member of the family.  But, much like we ask how
the snow plow driver gets to work \cite{Pratchett:uw}, we must ask how
these libraries are based are initially compiled.  Library-based tools
are useless for the discovery of new families, and cannot be applied
to a newly sequenced genome for which such a library has not been
compiled.

@@

To solve this problem we turn to {\it de novo} TE identification
tools, identifying TEs using only the genome
sequence information.  A number of such tools are discussed in the
literature.   RECON and PILER are based on
self-alignment, using WU-BLAST and LASTZ for the alignment
\cite{Bao:2002,Edgar:2005p2365,Lopez:2003td,Harris:2007uf}.  RECON show good sensitivity but is
computationally intensive and infeasible for use on whole genomes
(requiring 60 hours for 18Mb rice genome in a 2013 study), while
PILER achieves a good runtime with very low sensitivity
\cite{Jiang:2013jt}.  ReAS \cite{Li:2005he} and RepeatScout
\cite{Price:2005p1247} are based on $k$-mer searches, with the
earlier showing less sensitivity than RECON \cite{Jiang:2013jt}.
RepeatGluer \cite{Pevzner:2004p3157,Zhi:2006p3199} is
based on a variation of DeBrujin graphs, which allows for a
decomposition of TE families into domains, but is very, very
computationally expensive.  In Saha {\it et al.} the authors perform
an extensive comparison of the tools, and conclude that RepeatScout is
the best tool overall for assembled genomes, while ReAS the best when
dealing with unassembled sequence fragments \cite{Saha:2008dm}.

\subsection*{Elementary Repeats}
Another line of development, first proposed by Zheng and Lonardi
\cite{Zheng:2005bl}, involves the use of {\it elementary repeats}.
Similar to the RepeatGluer domains, elementary repeats are
decompositions of TEs into basic building blocks.  Identification of
these building blocks are sufficient for the purpose
of masking, and can be assembled into Transposable Elements for those
interested in TEs themselves.

While it is notoriously difficult to mathematically model transposable
elements \cite{Bao:2002}, elementary repeats are more conducive to a
formal description.  For a given genome, a nucleotide sequence $r$ is
an elementary repeat if: {\bf (1)} It is of at least length $l$ (the
length requirement); {\bf (2)} there are at least $f$ copies of $r$
appear (the frequency requirement) {\bf (3)} there is no proper
substring of $r$ of length $\leq l$ that appears in the genome
independently of $r$ (the minimality requirement); {\bf (4)} $r$ is a
maximal string w.r.t (1-3) (the maximality requirement).  Having
proposed this definition, Zheng and Lonardi developed an
identification algorithm that had a runtime quadratic in the genome
size. This was refined to linear time by He and also by Huo {\it et
  al.}  \cite{He:2006gpa,Huo:2009hoa} based on variations of suffix
tree approaches, but are limited in their ability to handle sequence
variation.  As we are looking a genome size inputs, and TEs
transposable elements naturally suffer from copy mistakes and
accumulate instance-specific base substitutions over time, this is a
significant limitation.

\subsection*{RAIDER}
\label{RAIDERSec}
It was with the objective of creating a linear time identification
algorithm that could handle variation through use of
PatternHunter-like spaced seeds that we developed the prototype RIADER
\cite{Li:2004wl}.  A rough implementation was first presented in
Figueroa {\it et al.}, with more details in the Figueroa masters
thesis \cite{Figueroa:2014uk,Figueroa:2013cz}.  RAIDER was built along an
alternate, but equivalent definition of elementary repeats based on
$l$-mers (sequences of length $l$).  Specifically, it was observed
that the minimality condition could be rewritten as: 
There is no $l$-mer contained within elementary repeat $r$ that appears in the genome
more times than $r$.  From there we make four core observations that form
the basis for the RAIDER algorithm:
\begin{enumerate}
\item An $l$-mer cannot belong to two different elementary repeats.
\item Any $l$-mer in the genome that occurs $f$ or more times is
  either an elementary repeat or belongs to one.
\item Any two $l$-mers belonging to an elementary repeat must appear
  the same number of times in the genome. 
\item If two sequences in the genome that are {\it maximally
  identical} (that is, cannot be extended in either direction and still
  be identical), these sequences cannot belong to a larger
  elementary repeat.
\end{enumerate}
By ``belong'' we mean ``is a substring of''; the reason for the
terminology will be explained shortly.  For discussion and proof, see
the Figueroa Thesis \cite{Figueroa:2013cz}.

@@

PatternHunter, a very successful augmentation to
BLAST \cite{Li:2004wl,Altschul:1997p843}, is based on the notion of {\it
  spaced seeds}: improving the sensitivity of string matching based
algorthms by allowing wild-cards in the match.  That is, instead of
requiring two strings matching in 12 consecutive characters, we might
instead require two six-character exact matches seperated by one base
(represented by the {\it seed pattern} 11111101111111), or perhaps
three consecutive four-character exact matches sperated by two bases
each (1111001111001111). It has been demonstrated that certain seed
patterns can idunce significant imporvements in BLAST sensitivity with
out time penalty, though what makes a good pattern is not well understood.

RAIDER was designed with the intent of employing the spaced seed
strategy, but this was only implemented heuristically for the Figuera
{\it et al.} paper \cite{Figueroa:2013cz} -- serving primarily as a
proof-of-concept.  Since its release we have develoed a formal model
of elementary repeats that incorporated spaced seeds, and from that
developed phRAIDER (PatternHunter-based RAIDER).  phRAIDER is a fast
tool for the identification and making of transposible elements in
both assembled and unassembled genomes aht outpreforms RptScout and
other establihsed tools. Code is free available under the Gnu GPL
lisence (v. 3) and  may be obtained [NEED WEB ADDRESS].  In our Methods section
we present the new model that allows us to extend RAIDER to correctly
use spaced seeds, and briefly outline the algorithm (with more details
provided in the supplementary materials), with a analysis of phRAIDER
performance in our Results.

@@

Our goal is to redefine the concept of transposbile elements to
accomodate a spaced seed strategy.  In the following section we will describe our
identification algorithm, and then describe the success in using it to
mase transposable elements. But we will start here with a brief
outline of our theoretical model (with a more detailed description in
the appendix).

\subsection*{phRAIDER}

In order to adapt the RAIDER algorithm to the new definition, we
require observations parallel to those on which RAIDER was built.  As
it turns out, its easy to adopt those.  Consider, for example, the
first observation: ``An $l$-mer cannot belong to two different
elementary repeats.''  We have already seen a violation, but can
eliminate this by changing the definition of {\it belongs}:
\begin{definition}
  Given a elementary repeat descriptor $r$ for seed $s$ with length
  $l$, we say a length $w$ substring \underline{belongs} to $r$ if it
  matches a member of the generalized decomposition of $r$ (w.r.t. to
  $s$).
\end{definition}
We notice in the above example, the string $AAAAA$ belongs to the
generalized decomposition, hence could not appear in both elementary
repeats.  But the string $AAAAT$ does not match any string in the
generalized decomposition of $r$, and hence was free to belong to the
descriptor of a ``distinct'' family.  See the Supplementary Materials
for a proof of all the core observations.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
\section*{Background}
Transposable Elements (TEs) are genomic sequences thathad the capacity
to insert copies of themselves into other genomic locations, resulting
in homologous families of sequences spread across the genome.  Present
in almost every higher order genome (covering as much as 45\% of the
human genome and 90\% of the maize genome
\cite{Venter:2001p92,SanMiguel:1996wa}), TEs have proved an important
source of data in numerous studies of genomic structure (e.g.
\cite{Arndt:2005p279,Karro:2008p123,Mugal:2009p581,Hardison:2003p97}).
But given their prevalence, it is important for those studying other
aspect the genome to have TEs masked out -- their bases replaced by N
to allow for easy identification and filtering.  Failure to filter can
reek havoc with genomic analysis tools.  For example, unfiltered TEs can trigger
huge numbers of false positives in automated gene finding tool
\cite{Jiang:2013jt}, as well as inflate tool runtime.

The best tools for repeat identification are RepeatMasker and nHMMER
\cite{RepeatMaskerOpen:XkNxXSd7,Wheeler:2013gj}, but both employ 
library-based search strategies using a pre-compiled
description of sequences in the family (e.g. a ancestral sequence for
BLASTing, or a profile HMM).  But, much like we ask how
the snow plow driver gets to work \cite{Pratchett:uw}, we must ask how
these libraries are compiled.  Library-based tools
are useless for the discovery of new families, and this are
challanging to use on newly sequenced genomes.

@@

To solve this problem we turn to {\it de novo} TE identification
tools, identifying TEs using only the genome sequence information.  A
number of such tools are discussed in the literature.  RECON uses
WU-BLAST and PILER using LASTZ to compute self-alignments
\cite{Bao:2002,Edgar:2005p2365,Lopez:2003td,Harris:2007uf}.  RECON
show good sensitivity but is computationally intensive and infeasible
for use on whole genomes (requiring 60 hours for 18Mb rice genome in a
2013 study), while PILER achieves a good runtime with very low
sensitivity .  ReAS and RepeatScout \cite{Price:2005p1247} are based
on $k$-mer searches, with the earlier showing less sensitivity than
RECON \cite{Li:2005he, Price:2005p1247,Jiang:2013jt}.  RepeatGluer
\cite{Pevzner:2004p3157,Zhi:2006p3199} is based on a variation of
DeBrujin graphs, which allows for a decomposition of TE families into
domains, but is very, very computationally expensive.  In Saha {\it et
  al.} the authors perform an extensive comparison of the tools, and
conclude that RepeatScout is the best tool overall for assembled
genomes, while ReAS the best when dealing with unassembled sequence
fragments \cite{Saha:2008dm}.

\subsection*{Elementary Repeats}
Zheng and Lonardi approached the {\it de novo} identification problem using {\it elementary repeats}
\cite{Zheng:2005bl}.
Similar to the RepeatGluer domains, elementary repeats are
decompositions of TEs into basic building blocks.  Identification of
these building blocks are sufficient for the purpose
of masking, and can be assembled into Transposable Elements for those
interested in TEs themselves.

While it is notoriously difficult to mathematically model transposable
elements \cite{Bao:2002}, elementary repeats are more conducive to a
formal description.  For a given genome, a nucleotide sequence $r$ is
an elementary repeat if: (1) It is of at least length $l$ (the length
requirement); (2) there are at least $f$ copies of $r$ appear (the
frequency requirement) (3) there is no proper substring of $r$ of
length $\leq l$ that appears in the genome independently of $r$ (the
minimality requirement); (4) $r$ is a maximal string w.r.t (1-3) (the
maximality requirement).  Having proposed this definition, Zheng and
Lonardi developed an identification algorithm that had a runtime
quadratic in the query sequence size \cite{Zheng:2005bl}. This was
refined to linear time by He and also by Huo {\it et al.}
\cite{He:2006gpa,Huo:2009hoa} based on variations of suffix tree
approaches, but these appraoches are limited in their ability to handle sequence
variation.  As we are looking a genome size inputs, and TEs
transposable elements naturally suffer from copy mistakes and
accumulate instance-specific base substitutions over time, this is a
significant limitation.

\subsection*{RAIDER}
\label{RAIDERSec}
It was with the objective of creating a linear time identification
algorithm that could handle variation through use of
PatternHunter-like spaced seeds that we developed the prototype RIADER
\cite{Li:2004wl}.  A rough implementation was first presented in
Figueroa {\it et al.}, with more details in the Figueroa masters
thesis \cite{Figueroa:2014uk,Figueroa:2013cz}.  RAIDER was built along
an alternate, but equivalent definition of elementary repeats based on
$l$-mers (sequences of length $l$).  Specifically, it was observed
that the minimality condition could be rewritten as: There is no
$l$-mer contained within elementary repeat $r$ that appears in the
genome more times than $r$.  From there we make four core observations
that form the basis for the RAIDER algorithm: (1) An $l$-mer cannot
belong to two different elementary repeats; (2) Any $l$-mer in the
genome that occurs $f$ or more times is either an elementary repeat or
belongs to one; (3) any two $l$-mers belonging to an elementary repeat
must appear the same number of times in the genome; (4) If two
sequences in the genome that are {\it maximally identical} (that is,
cannot be extended in either direction and still be identical), these
sequences cannot belong to a larger elementary repeat.  (By ``belong''
we mean ``is a substring of'' -- a definition we will be generalizing
shortly.)  For discussion and proof, see the Figueroa Thesis
\cite{Figueroa:2013cz}.

@@

PatternHunter, a very successful augmentation to BLAST
\cite{Li:2004wl,Altschul:1997p843}, is based on the notion of {\it
  spaced seeds}: improving the sensitivity of string matching based
algorthms by allowing wild-cards in the match.  That is, instead of
requiring two strings matching in 12 consecutive characters, we might
instead require two six-character exact matches seperated by one base
which may or may not match (represented by the {\it seed pattern}
11111101111111), or perhaps three consecutive four-character exact
matches sperated by two bases each (1111001111001111). It has been
demonstrated that certain seed patterns can indunce significant
imporvements in BLAST sensitivity with out time penalty, though what
makes a good pattern is not well understood.

RAIDER was designed with the intent of employing the spaced seed
strategy, but this was only implemented heuristically for the Figuera
{\it et al.} paper \cite{Figueroa:2013cz} -- serving primarily as a
proof-of-concept.  Since its release we have develoed a formal model
of elementary repeats that incorporated spaced seeds, and from that
developed phRAIDER (PatternHunter-based RAIDER).  phRAIDER is a fast
tool for the identification and making of transposible elements in
both assembled and unassembled genomes aht outpreforms RptScout and
other establihsed tools. Code is free available under the Gnu GPL
lisence (v. 3) and  may be obtained [NEED WEB ADDRESS].

In the following we will present a new theoretical model for {\it
  seeded elementary repeats}, followed by an algorithm for identifying
them.  In the results section we will show the result of using that
algorithm for masking transp

In our Methods section
we present the new model that allows us to extend RAIDER to correctly
use spaced seeds, and briefly outline the algorithm (with more details
provided in the supplementary materials), with a analysis of phRAIDER
performance in our Results.

@@

Our goal is to redefine the concept of transposbile elements to
accomodate a spaced seed strategy.  In the next section we will describe our
identification algorithm, and then quantified its success in 
masking transposable elements. But we will start here with a brief
outline of our theoretical model (with a more detailed description in
the appendix).

We first need to redfine out terminimology regarding elementary
repeats.  Under the Z\&L definition, all instances of one elementary
repeat have the exact same sequence, and hence can be descrbied by a
single string.  As we are allowing for variation, we will describe out
instance set with a {\it sequence descriptor} consisting of bases
letters and the wild-card character * (e.g. AAC*GG would describe a
set of sequences with starting with an AAC, ending with a GG, and
having any character in between).  Given a binary string $s$
representing a spaced seed, we say a sequence descriptor $r$ is
{\it consistent} with $s$ if we can align $s$ to a substring of $r$ such
that every $*$ in $r$ aligns with a $0$ character in $s$. (Hence
$ACG**T*A$ is consistent with the seed $11001$, but not $11011$.)

Given a sequence descriptor $r$, we can {\it decompose} $r$ with
respect to $s$ by taking every length $|s|$ substring of $r$ that is consistant
with $s$, replacing all letters of $r$ that match to a 0 in $s$ with a
$*$, and creating a set from the results.  (Hence for $s = 11011$ and
$r=AA*CCGTT$, the decomposition would be $\{AA*CC, CC*TT\}$.)  We say
$s$ {\it covers} $r$ is every base in $r$ is contained in at least on
string in the decompositio.  (Hence the previous $s$ does not cover $AAA*CC$, as
the first base in not in any of the strings of the decomposition.)

We can now modif the previous definition of elementary repeats as
follows:
\begin{definition} Given a genomic sequence $G$, an integer $f$, and a
  spaced seed $s$, a sequence descriptor $r$ describes an elementary
  repeat if it meets the four (moified) requirements of an elementary repeat:
\begin{itemize}
\item \underline{Structure requriement}: $s$ covers $r$.
\item \underline{Frequency requirement}: There are at least $f$
  substrings of $G$ that match $r$.
\item \underline{Minimality requirement:} For every string $t$ in the
  decomposition of $r$ w.r.t $s$, the number of occurences of $t$ in
  $G$ is equal to the number of occurneces of $r$ in the genome.
\item \underline{Maximality requirement:} There is no sequence
  descriptor $r'$ of $r$ that contains $r$ as a proper substring and
  satisfies conditions 1-1.
\end{itemize}
\end{definition}

\begin{theorem} When the seed $s$ has no 0 characters, this
    definition of elementary repeats is equivilent to the Z\&L
    definition.
\end{theorem}
